---
title: "Numeric Data"
---

```{r}
#| include: false
options(warn = -1)
```

::: callout-note
## Abstract

This section reviews how to access data that is primarily numeric/quantitative in nature, but from a different source and of a different nature than the data typically used by social scientists. Example datasets include international or large-scale assessments (e.g., PISA, NAEP, IPEDS) and data from digital technologies (e.g., log-trace data from the Open University Learning Analytics Dataset (OULAD)).
:::

## 4.1 Overview

In social science research, data is traditionally sourced from small-scale surveys, experiments, or qualitative studies. However, the rise of big data offers researchers opportunities to explore numeric and quantitative datasets of unprecedented scale and variety. This chapter discusses how to access and analyze large-scale datasets like international assessments (e.g., PISA, NAEP) and digital log-trace data (e.g., Open University Learning Analytics Dataset (OULAD)). These secondary data sources enable novel research questions and methods, particularly when paired with machine learning and statistical modeling approaches.

## 4.2 Accessing Big Data (Broadening the Horizon)

### 4.2.1 Big Data

#### Accessing PISA Data

The Programme for International Student Assessment (PISA) is a widely used dataset for large-scale educational research. It assesses 15-year-old students' knowledge and skills in reading, mathematics, and science across multiple countries. Researchers can access PISA data through various methods:

##### 1. Direct Download from the Official Website

The OECD provides direct access to PISA data files via its official website. Researchers can download data for specific years and cycles. Data files are typically provided in `.csv` or `.sav` (SPSS) formats, along with detailed documentation.

-   **Steps to Access PISA Data from the OECD Website:**
    1.  Visit the [OECD PISA website](https://www.oecd.org/pisa/).
    2.  Navigate to the "Data" section.
    3.  Select the desired assessment year (e.g., 2022).
    4.  Download the data and accompanying codebooks.

##### 2. Using the `OECD` R Package

The `OECD` R package provides a direct interface to download and explore datasets published by the OECD, including PISA.

-   **Steps to Use the `OECD` Package:**
    1.  Install and load the `OECD` package.
    2.  Use the `get_datasets()` and `get_dataset()` functions (note: the API may vary by package version).

```{r}
#| label: pisa-oecd-example
#| eval: false
# Install and load the OECD package
# install.packages("OECD")
library(OECD)

# Fetch PISA data for the 2022 cycle (Example)
pisa_data <- get_dataset("PISA_2022")

# Display a summary of the data
summary(pisa_data)
```

##### 3. Using the `EdSurvey` R Package

The `EdSurvey` package is designed specifically for analyzing large-scale assessment data, including PISA. It allows for complex statistical modeling and supports handling weights and replicate weights used in PISA.

-   **Steps to Use the `EdSurvey` Package:**
    1.  Install and load the `EdSurvey` package.
    2.  Download the PISA data from the OECD website and provide the path to the `.sav` files.
    3.  Load the data into R using `readPISA()`.

```{r}
#| label: pisa-edsurvey-example
#| eval: false
# Install and load the EdSurvey package
# install.packages("EdSurvey")
library(EdSurvey)

# Read PISA data from a local file
pisa_data <- readPISA("path/to/PISA2022Student.sav")

# Display the structure of the dataset
str(pisa_data)
```

##### Comparison of Methods

| Method | Advantages | Disadvantages |
|----------------|-----------------------------|---------------------------|
| **Direct Download** | Full access to all raw data and documentation. | Requires manual processing and cleaning. |
| **`OECD` Package** | Easy to use for downloading specific datasets. | Limited to OECD-published formats. |
| **`EdSurvey` Package** | Supports advanced statistical analysis and weights. | Requires additional setup and dependencies. |

#### Accessing IPEDS Data

The Integrated Postsecondary Education Data System (IPEDS) is a comprehensive source of data on U.S. colleges, universities, and technical and vocational institutions. It provides data on enrollments, completions, graduation rates, faculty, finances, and more. Researchers and policymakers widely use IPEDS data to analyze trends in higher education.

There are several ways to access IPEDS data, depending on the user's needs and technical proficiency.

##### 1. Direct Download from the NCES Website

The most straightforward way to access IPEDS data is by downloading it directly from the National Center for Education Statistics (NCES) website.

-   **Steps to Access IPEDS Data:**
    1.  Visit the [IPEDS Data Center](https://nces.ed.gov/ipeds/).
    2.  Click on "Use the Data" and navigate to the "Download IPEDS Data Files" section.
    3.  Select the desired data year and survey component (e.g., Fall Enrollment, Graduation Rates).
    4.  Download the data files, typically provided in `.csv` or `.xls` format, along with accompanying codebooks.

##### 2. Using the `ipeds` R Package

The `ipeds` R package simplifies downloading and analyzing IPEDS data directly from R by connecting to the NCES data repository.

-   **Steps to Use the `ipeds` Package:**
    1.  Install and load the `ipeds` package.
    2.  Use the `download_ipeds()` function to fetch data for specific survey components and years.

```{r}
#| label: ipeds-package-example
#| eval: false
# Install and load the ipeds package
# install.packages("ipeds")
library(ipeds)

# Download IPEDS data for completions in 2021
ipeds_data <- download_ipeds("C", year = 2021)

# View the structure of the downloaded data
str(ipeds_data)
```

##### 3. Using the `tidycensus` R Package

The `tidycensus` package, while primarily designed for Census data, can access specific IPEDS data linked to educational institutions.

-   **Steps to Use the `tidycensus` Package:**
    1.  Install and load the `tidycensus` package.
    2.  Set up a Census API key to access the data.
    3.  Query IPEDS data for specific institution-level information.

```{r}
#| label: tidycensus-example
#| eval: false
# Install and load the tidycensus package
# install.packages("tidycensus")
library(tidycensus)

# Set Census API key (replace with your actual key)
census_api_key("your_census_api_key")

# Fetch IPEDS-related data (e.g., institution information)
ipeds_institutions <- get_acs(
  geography = "place",
  variables = "B14002_003",
  year = 2021,
  survey = "acs5"
)

# View the first few rows
head(ipeds_institutions)
```

##### 4. Using Online Tools

IPEDS provides several online tools for querying and visualizing data without requiring programming skills.

-   **Common Tools:**
    -   **IPEDS Data Explorer**: Enables users to query and export customized datasets.
    -   **Trend Generator**: Allows users to visualize trends in key metrics over time.
    -   **IPEDS Use the Data**: Simplified tool for accessing pre-compiled datasets.
-   **Steps to Use the IPEDS Data Explorer:**
    1.  Visit the IPEDS Data Explorer.
    2.  Select variables of interest, such as institution type, enrollment size, or location.
    3.  Filter results by years, institution categories, or other criteria.
    4.  Export the results as a `.csv` or `.xlsx` file.

##### Comparison of Methods

| Method | Advantages | Disadvantages |
|-----------------|----------------------------|----------------------------|
| **Direct Download** | Full access to raw data and documentation. | Requires manual data preparation and cleaning. |
| **`ipeds` Package** | Automated access to specific components. | Limited flexibility for customized queries. |
| **`tidycensus` Package** | Allows integration with Census and ACS data. | Requires API setup and advanced R skills. |
| **Online Tools** | User-friendly and suitable for non-coders. | Limited to predefined queries and exports. |

#### Accessing Open University Learning Analytics Dataset (OULAD)

The **Open University Learning Analytics Dataset (OULAD)** is a publicly available dataset designed to support research in educational data mining and learning analytics. It includes student demographics, module information, interactions with the virtual learning environment (VLE), and assessment scores.

-   **Steps to Access OULAD Data:**
    1.  Visit the [OULAD Repository](https://archive.ics.uci.edu/dataset/349/open+university+learning+analytics+dataset).
    2.  Download the dataset as a `.zip` file.
    3.  Extract the `.zip` file to a local directory.

The dataset contains multiple CSV files: - `studentInfo.csv`: Student demographics and performance data. - `studentVle.csv`: Interactions with the VLE. - `vle.csv`: Details of learning resources. - `studentAssessment.csv`: Assessment scores.

##### Loading OULAD Data in R

Once the data is downloaded and extracted, follow these steps to load and access it in R:

```{r}
#| label: load-oulad-example
#| eval: false
# Install necessary packages
# install.packages(c("readr", "dplyr"))
library(readr)
library(dplyr)

# Define the path to the OULAD data
data_path <- "path/to/OULAD/"

# Load individual CSV files
student_info <- read_csv(file.path(data_path, "studentInfo.csv"))
student_vle <- read_csv(file.path(data_path, "studentVle.csv"))
vle <- read_csv(file.path(data_path, "vle.csv"))
student_assessment <- read_csv(file.path(data_path, "studentAssessment.csv"))

# Inspect the structure and contents
head(student_info)
str(student_vle)
```

## 4.3 Logistic Regression & Machine Learning

### 4.3.1 Purpose + Case

#### Purpose

Logistic regression is a supervised learning technique widely used for binary classification tasks. It models the probability of an event occurring (e.g., success vs. failure) based on a set of predictor variables. Logistic regression is particularly effective in educational research for predicting outcomes such as retention, enrollment, or graduation rates.

#### Case Study: Predicting Graduation Rates

This case study is based on IPEDS data and inspired by Zong and Davis (2022). We predict graduation rates as a binary outcome (`good_grad_rate`) using institutional features such as total enrollment, admission rate, tuition fees, and average instructional staff salary.

### 4.3.2 Sample Research Questions

-   **RQ A**: What institutional factors are associated with high graduation rates in U.S. four-year universities?
-   **RQ B**: How accurately can we predict high graduation rates using institutional features with supervised machine learning?

### 4.3.3 Analysis

#### Step 1: Load Required Packages

We load necessary R packages for data wrangling, cleaning, and modeling.

```{r}
#| label: load-libraries-logistic
# Load necessary libraries for data cleaning, wrangling, and modeling
library(tidyverse)  # For data manipulation and visualization
library(tidymodels) # For machine learning workflows
library(janitor)    # For cleaning variable names
```

#### Step 2: Load and Clean Data

We read the IPEDS dataset and clean column names for easier handling.

```{r}
#| label: load-data-logistic
# Read in IPEDS data from CSV file
ipeds <- read_csv("data/ipeds-all-title-9-2022-data.csv")

# Clean column names for consistency and usability
ipeds <- ipeds |>
  clean_names()
```

#### Step 3: Data Wrangling

Select relevant variables, filter the dataset, and create the dependent variable `good_grad_rate`.

```{r}
#| label: wrangle-data-logistic
# Select and rename key variables; filter relevant institutions
ipeds <- ipeds |>
  select(
    name = institution_name,                  # Institution name
    total_enroll = drvef2022_total_enrollment, # Total enrollment
    pct_admitted = drvadm2022_percent_admitted_total, # Admission percentage
    tuition_fees = drvic2022_tuition_and_fees_2021_22, # Tuition fees
    grad_rate = drvgr2022_graduation_rate_total_cohort, # Graduation rate
    percent_fin_aid = sfa2122_percent_of_full_time_first_time_undergraduates_awarded_any_financial_aid, # Financial aid
    avg_salary = drvhr2022_average_salary_equated_to_9_months_of_full_time_instructional_staff_all_ranks # Staff salary
  ) |>
  filter(!is.na(grad_rate)) |> # Remove rows with missing graduation rates
  mutate(
    # Create binary dependent variable for high graduation rates (median split or threshold)
    good_grad_rate = if_else(grad_rate > 62, 1, 0),
    good_grad_rate = as.factor(good_grad_rate) # Convert to factor for classification
  )
```

#### Step 4: Exploratory Data Analysis (EDA)

Visualize the distribution of the graduation rate.

```{r}
#| label: eda-logistic
# Plot a histogram of graduation rates
ipeds |>
  ggplot(aes(x = grad_rate)) +
  geom_histogram(bins = 20, fill = "blue", color = "white") +
  theme_minimal() +
  labs(
    title = "Distribution of Graduation Rates",
    x = "Graduation Rate",
    y = "Frequency"
  )
```

#### Step 5: Logistic Regression Model

Fit a logistic regression model to predict high graduation rates.

```{r}
#| label: model-logistic
# Fit logistic regression model
m1 <- glm(
  good_grad_rate ~ total_enroll + pct_admitted + tuition_fees + percent_fin_aid + avg_salary,
  data = ipeds,
  family = "binomial" # Specify logistic regression for binary outcome
)

# View model summary
summary(m1)
```

#### Step 6: Supervised Machine Learning Workflow

Use the `tidymodels` framework to build a machine learning model.

```{r}
#| label: ml-workflow-logistic
# Define recipe for the model (preprocessing steps)
my_rec <- recipe(good_grad_rate ~ total_enroll + pct_admitted + tuition_fees + percent_fin_aid + avg_salary, data = ipeds)

# Specify logistic regression model with tidymodels
my_mod <- logistic_reg() |>
  set_engine("glm") |>         # Use glm engine for logistic regression
  set_mode("classification")    # Specify binary classification task

# Create workflow to connect recipe and model
my_wf <- workflow() |>
  add_recipe(my_rec) |>
  add_model(my_mod)

# Fit the logistic regression model
fit_model <- fit(my_wf, ipeds)

# Generate predictions on the dataset
predictions <- predict(fit_model, ipeds) |>
  bind_cols(ipeds) # Combine predictions with original data

# Calculate and display accuracy
my_accuracy <- predictions |>
  metrics(truth = good_grad_rate, estimate = .pred_class) |>
  filter(.metric == "accuracy")

my_accuracy
```

### 4.3.4 Results and Discussions

#### Logistic Regression Model (RQ A)

The logistic regression model was fitted to predict whether a university achieves a "good" graduation rate (i.e., graduation rate > 62%) based on several institutional features. The model output is summarized below:

-   **Coefficients & Significance**:
    -   **total_enroll**: Estimate = 3.35e-05, *z* = 4.251, p = 2.13e-05\
        *Interpretation*: Simply put, bigger schools tend to graduate more students. It seems that larger institutions might have more structured support systems in place to keep students on track.
    -   **pct_admitted**: Estimate = -1.407e-02, *z* = -3.997, p = 6.40e-05\
        *Interpretation*: Here is a classic "selectivity" story. The more competitive a school is (admitting fewer students), the higher its graduation rate tends to be. This tells us that who you let in matters a lot for the final outcome.
    -   **tuition_fees**: Estimate = 6.952e-05, *z* = 14.003, p < 2e-16\
        *Interpretation*: This is the heavy hitter. Schools that charge higher tuition—and presumably invest that money back into student support, faculty, and facilities—enjoy significantly better graduation outcomes. It's a strong signal that money matters, but it's likely *how that money is used* that makes the difference.
    -   **percent_fin_aid**: Estimate = -2.960e-02, *z* = -5.237, p = 1.64e-07\
        *Interpretation*: This one is a bit counter-intuitive at first glance. A higher percentage of students receiving aid is linked to *lower* graduation rates. But this doesn't mean financial aid is bad! It likely reflects the student population these institutions serve. Schools with more students needing aid often face systemic challenges—like balancing work and study—that can hinder timely graduation.
    -   **avg_salary**: Estimate = 2.996e-05, *z* = 7.740, p = 9.91e-15\
        *Interpretation*: You get what you pay for. Schools paying their faculty well are associated with higher graduation rates, suggesting that attracting and retaining talented instructors directly impacts student success.
-   **Model Fit Statistics**:
    -   **Null Deviance**: 2277 (on 1706 degrees of freedom)
    -   **Residual Deviance**: 1632 (on 1701 degrees of freedom)
    -   **AIC**: 1644
    -   *Note*: 3621 observations were deleted due to missing values.

Overall, the regression model demonstrates that several institutional factors are statistically significant predictors of graduation rates. In particular, tuition fees and average staff salary have a strong positive effect, while admission selectivity and financial aid show negative associations.

#### Supervised ML Workflow Results (RQ B)

Using the tidymodels framework, we built a logistic regression model as part of a supervised machine learning workflow. The performance metric obtained is as follows:

-   **Accuracy**: 80.02%

This tells us that based on just five or six simple institutional metrics, we can correctly classify whether a university is "high performing" or not with about **80% accuracy**. That's pretty powerful! It means these numbers aren't just random noise—they tell a consistent story about what makes a university tick.

#### Overall Discussion

-   **Similarities between Approaches**:
    -   Both the traditional logistic regression and the tidymodels workflow identified key predictors that influence graduation rates, such as total enrollment, admission percentage, tuition fees, financial aid percentage, and average staff salary.
    -   Each approach provides valuable insights: the regression model offers detailed coefficient estimates and significance levels, while the tidymodels workflow emphasizes predictive accuracy.
-   **Differences between Approaches**:
    -   **Interpretability vs. Predictive Performance**: The logistic regression output delivers interpretability through its coefficients and p-values, allowing us to understand the direction and magnitude of the relationships. In contrast, the supervised ML workflow focuses on achieving a robust predictive performance, evidenced by an 80% accuracy.
    -   **Handling of Data**: The traditional regression model summarizes the relationship between variables, whereas the ML workflow integrates data pre-processing, modeling, and validation into a cohesive framework.

In summary, our analyses indicate that institutional factors, particularly tuition fees and staff salaries, play a significant role in predicting graduation outcomes. The supervised ML approach, with an accuracy of around 80%, confirms the model's practical utility in classifying institutions based on graduation performance. Both methods complement each other, providing a comprehensive understanding of the underlying dynamics that drive graduation rates in higher education.

## 4.4 Random Forests & Learning Analytics

In this section, we explore a more sophisticated supervised learning approach—Random Forests—to model student interaction data from the Open University Learning Analytics Dataset (OULAD).

### 4.4.1 Purpose + Case

#### Purpose

Random Forests is an ensemble learning method that builds multiple decision trees and aggregates their results to improve prediction accuracy and control overfitting. It is particularly well-suited for complex, high-dimensional data such as student interaction (clickstream) data.

#### Case Study: Predicting Student Success with VLE Data

Inspired by research on digital trace data, this case study uses interactions data from OULAD. We focus on predicting whether a student will pass a course based on engineered features derived from clickstream data, such as total clicks, mean clicks, and linear trends over time.

### 4.4.2 Sample Research Questions

-   **RQ1**: How accurately can a random forest model predict whether a student will pass a course using interaction data from OULAD?
-   **RQ2**: Which interaction-based features (e.g., total clicks, clickstream slope) are most important in predicting student outcomes?
-   **RQ3**: How does the use of cross-validation influence the stability and generalizability of the model?

### 4.4.3 Analysis

#### Step 1: Load Required Packages

```{r}
#| label: load-libraries-rf
# Load necessary libraries for data manipulation and modeling
library(tidyverse)      # Data wrangling and visualization
library(janitor)        # Cleaning variable names
library(tidymodels)     # Modeling workflow
library(ranger)         # Random forest implementation
library(vip)            # Variable importance plots
```

#### Step 2: Load and Prepare Data

We load and join multiple OULAD files to create a complete modeling dataset.

```{r}
#| label: load-data-rf
# Load the interactions data and student performance data
interactions <- read_csv("data/oulad-interactions-filtered.csv")
students_and_assessments <- read_csv("data/oulad-students-and-assessments.csv")
assessments <- read_csv("data/oulad-assessments.csv")

# Create cutoff dates based on first 25% of assessment submissions
code_module_dates <- assessments |> 
    group_by(code_module, code_presentation) |> 
    summarize(quantile_cutoff_date = quantile(date_submitted, probs = 0.25, na.rm = TRUE), .groups = 'drop')

# Filter interactions to include only those before the cutoff
interactions_filtered <- interactions |> 
    left_join(code_module_dates, by = c("code_module", "code_presentation"), suffix = c(".x", "")) |> 
    filter(date < quantile_cutoff_date) |> 
    select(-any_of("quantile_cutoff_date.x"))

# Feature Engineering: Summary statistics of clicks
interactions_summarized <- interactions_filtered |> 
    group_by(id_student, code_module, code_presentation) |> 
    summarize(
      sum_clicks = sum(sum_click, na.rm = TRUE),
      sd_clicks = sd(sum_click, na.rm = TRUE), 
      mean_clicks = mean(sum_click, na.rm = TRUE),
      .groups = "drop"
    )

# Feature Engineering: Clickstream slopes (trends)
fit_slope <- function(df) {
    if(nrow(df) < 2) return(tibble(term = c("(Intercept)", "date"), estimate = c(NA, NA)))
    model <- lm(sum_click ~ date, data = df)
    broom::tidy(model) |> select(term, estimate)
}

interactions_slopes <- interactions_filtered |>
    group_by(id_student, code_module, code_presentation) |>
    nest() |>
    mutate(model = map(data, fit_slope)) |>
    select(-data) |>
    unnest(model) |>
    pivot_wider(names_from = term, values_from = estimate) |>
    rename(intercept = `(Intercept)`, slope = date) |>
    mutate(across(where(is.numeric), \(x) round(x, 4)))

# Join all features
students_full <- students_and_assessments |> 
    left_join(interactions_summarized, by = c("id_student", "code_module", "code_presentation")) |> 
    left_join(interactions_slopes, by = c("id_student", "code_module", "code_presentation")) |> 
    mutate(pass = as.factor(pass))
```

#### Step 3: Define Model Recipe

```{r}
#| label: recipe-rf
my_rec2 <- recipe(pass ~ disability + date_registration + gender + code_module +
                     mean_weighted_score + sum_clicks + sd_clicks + mean_clicks + 
                     intercept + slope, 
                 data = students_full) |> 
    step_dummy(all_nominal_predictors()) |> 
    step_impute_knn(all_predictors()) |> 
    step_normalize(all_numeric_predictors())
```

#### Step 4: Specify Model and Workflow

```{r}
#| label: workflow-rf
# Specify random forest model with impurity-based importance
my_mod2 <- rand_forest() |> 
    set_engine("ranger", importance = "impurity") |> 
    set_mode("classification")

# Create workflow
my_wf2 <- workflow() |> 
    add_recipe(my_rec2) |> 
    add_model(my_mod2)
```

#### Step 5: Resampling and Model Fitting

```{r}
#| label: fit-resamples-rf
set.seed(20230712)
vfcv <- vfold_cv(data = students_full, v = 4, strata = pass)
class_metrics <- metric_set(accuracy, sensitivity, specificity, ppv, npv, kap)

# Fit using cross-validation
fitted_resamples <- fit_resamples(my_wf2, resamples = vfcv, metrics = class_metrics)
collect_metrics(fitted_resamples)
```

#### Step 6: Final Evaluation

```{r}
#| label: final-fit-rf
set.seed(20230712)
train_test_split <- initial_split(students_full, prop = 0.67, strata = pass)

# Final fit on training, evaluate on test
final_fit <- last_fit(my_wf2, train_test_split, metrics = class_metrics)
collect_metrics(final_fit)

# Variable Importance Plot
final_fit |> 
  extract_workflow() |> 
  extract_fit_parsnip() |> 
  vip(num_features = 10) +
  theme_minimal() +
  labs(title = "Top 10 Important Features in Random Forest Model")
```

### 4.4.4 Results and Discussions

#### Research Question 1 (RQ1):

*How accurately can a random forest model predict whether a student will pass a course using interactions data from OULAD?*

**Response:**\
Using 4-fold cross-validation, our random forest model yielded an average accuracy of approximately **67.0%** (mean accuracy from resamples: 0.670) with a Cohen's Kappa of **0.261**, suggesting moderate agreement beyond chance. When fitted on the entire training set and evaluated on the test set, the final model showed an accuracy of **66.5%** along with:
- **Sensitivity:** 81.5% – indicating the model correctly identifies a high proportion of students who pass.
- **Specificity:** 41.9% – suggesting that the model is less effective at correctly identifying students who do not pass.
- **Positive Predictive Value (PPV):** 69.7%
- **Negative Predictive Value (NPV):** 58.0%

The confusion matrix shows:
- True Negatives (TN): 5439
- False Negatives (FN): 1238
- False Positives (FP): 2370
- True Positives (TP): 1710

Overall, these metrics indicate that while the model performs well in detecting positive outcomes (high sensitivity), its lower specificity means that it tends to misclassify a relatively higher proportion of non-passing students. In plain English, our model is an **optimist**—it's great at spotting the winners, but it sometimes overlooks the stragglers. This is a crucial insight for educators: this tool is better used as an **early warning system for success** rather than a perfect detector of failure.

#### Research Question 2 (RQ2):

*Which interaction-based features are most important in predicting student outcomes?*

**Response:**\
The variable importance analysis, extracted from the final random forest model using the `vip()` function, highlights the following key predictors (with their respective importance scores):

-   **sum_clicks:** 1062.49 – This is the most influential feature, indicating that the total number of clicks (i.e., student engagement) in the VLE is a strong predictor of student success. In other words, "showing up" digitally matters just as much as showing up in person.
-   **mean_weighted_score:** 838.06 – Reflecting academic performance as measured by weighted assessment scores. Past performance is usually a strong predictor of future success, and this confirms it.
-   **mean_clicks:** 737.79, **slope:** 739.86, and **intercept:** 729.52 – These engineered features representing the central tendency and trend of click behavior further underline the importance of digital engagement patterns.
-   **date_registration:** 652.15 – The registration date also plays a significant role.
-   Other categorical variables (e.g., dummy-coded `disability`, `gender`, and `code_module` levels) generally show lower importance scores, with values typically under 80, indicating that while they do contribute, engagement and performance metrics dominate.

These results suggest that both the **intensity** and the **temporal trend** of student interactions with the learning environment are critical in predicting whether a student will pass. Specifically, the `slope` (trend) is a fascinating finding—it tells us that it's not just *how much* a student clicks, but *how their behavior changes* over time. A student whose engagement is increasing—even if they started slow—is often in a better position than one whose activity is tapering off.

#### Research Question 3 (RQ3):

*How does the use of cross-validation impact the stability and generalizability of the random forest model on interactions data?*

**Response:**\
The use of 4-fold cross-validation (via `vfold_cv`) allowed us to assess the model's performance across multiple subsets of the data, mitigating the risk of overfitting. The resampling results are relatively consistent (with accuracy around 67%, sensitivity at 81.6%, and specificity around 43.2%), which supports the model's robustness and generalizability. Although the final test set performance (accuracy of 66.5%) is slightly lower, the overall consistency of metrics across folds indicates that our model is stable when applied to unseen data. This consistency is key—it suggests that the patterns we've discovered aren't just a fluke; they likely hold true for other student populations beyond this dataset.

#### Overall Discussion

The random forest model built on interactions data from OULAD demonstrates decent predictive performance with an accuracy of approximately 66.5–67% and high sensitivity (around 81.5%), indicating strong capability in identifying students who will pass the course. However, the relatively low specificity (around 42%) suggests that there is room for improvement in correctly classifying students who are at risk of not passing.

The variable importance analysis underscores that engagement-related features—especially **sum_clicks** and features capturing the trend in interactions (slope, mean_clicks)—are the most influential predictors. This insight implies that the **digital footprint** of student engagement in the virtual learning environment is critical for predicting academic outcomes. Every click tells a story; ignoring them means missing valuable signals.

In summary, while our model performs robustly across cross-validation folds and provides actionable insights into key predictive features, the lower specificity points to the need for further refinement. Future work might explore additional feature engineering, alternative model tuning, or combining models to better balance sensitivity and specificity, ultimately supporting timely interventions in educational settings.
