---
title: "Colophon"
execute:
  freeze: false
---

This book was written in [Quarto](https://quarto.org/), analyzed with R, and rendered to HTML and PDF with XeLaTeX. Source files, data paths, package declarations, and version history are available in the book's [GitHub repository](https://github.com/data-edu/computational-social-science-cookbook).

## About the Authors

**Wei Wang** is a PhD Candidate in Learning, Design, and Technology at the University of Tennessee, Knoxville. His research focuses on generative AI in education, educational data science, and evaluation methods. Website: <https://wwang93.github.io/Vino-Wang-Website/>

**Mete Akcaoglu** is Professor of Instructional Technology and Chair of the Department of Learning, Teaching, and Human Development at Georgia Southern University. His research examines game-based learning, interest development, computational thinking, and online pedagogy, with a methodological focus on computational approaches to educational data. Website: <https://meteakcaoglu.com/>

**Joshua M. Rosenberg** is Haslam Family Professor and Associate Professor of STEM Education at the University of Tennessee, Knoxville. His research focuses on learning analytics, educational data science, and open, reproducible research in education. Website: <https://joshuamrosenberg.com/>

**Shaun Kellogg** is Senior Director of the Office of Research and Promising Practices at the North Carolina Department of Public Instruction. His work focuses on evidence-based decision-making, research-practice partnerships, and the use of data to improve educational outcomes. Website: <https://www.dpi.nc.gov/blog/2025/09/17/how-ncdpis-office-research-promising-practices-turns-evidence-action>

:::: {.content-visible when-format="html"}

## Reproducibility Details

This text-only update was rendered on September 29, 2026 with Quarto 1.10.18. The analysis outputs, build metadata, and R session information below are retained from the earlier edition.

```{r}
#| echo: false
#| message: false
#| warning: false

if (knitr::is_html_output()) {
  build_date <- format(Sys.Date(), "%B %d, %Y")
  quarto_version_file <- file.path(
    dirname(Sys.getenv("QUARTO_BIN_PATH")),
    "share",
    "version"
  )
  quarto_version <- if (file.exists(quarto_version_file)) {
    readLines(quarto_version_file, n = 1)
  } else {
    "Unavailable"
  }
  cat("Build date:", build_date, "\n")
  cat("Quarto version:", quarto_version, "\n")
}
```

::: {.callout-note collapse="true"}
## Full R session information

```{r}
#| echo: false
#| message: false
#| warning: false

if (knitr::is_html_output()) {
  print(sessionInfo())
}
```
:::

::::

::: {.content-visible when-format="pdf"}
Production details and source history for this edition are maintained in the repository.
:::
