# Section 3 **Multimodal Data (Images, Video, Audio)**

## 3.1 Overview

In this section, we will discuss how social scientists can look beyond conventional data types, and learn about capturing and analyzing multimodal data. 

## 3.2 Images

I am not sure if we need to talk about "accessing" or collecting image data as it is pretty much standard.

## 3.3 Analyzing Images

With the advent of Large-Language Models (LLMs) we can use their power to analyze images. In this section, we will focus on using one package that uses local LLMs (i.e., privacy) to analyze image files: [{kuzco}](https://github.com/frankiethull/kuzco).

### 3.3.1. Setting Up Kuzco

To use kuzco, you need to, first, install Ollama (a software that allows pulling and running local LLMs) and ollamar & ellmer packages.

You can install Ollama by downloading and installing the application from its provider's website. Basically the steps are:

1.  Download and install the [Ollama](https://ollama.com/) app.

-   [macOS](https://ollama.com/download/Ollama-darwin.zip)

-   [Windows preview](https://ollama.com/download/OllamaSetup.exe)

-   Linux: `curl -fsSL https://ollama.com/install.sh | sh`

-   [Docker image](https://hub.docker.com/r/ollama/ollama)

2.  Open/launch the Ollama app to start the local server.

After installing Ollama, you will then need to install ollamar and ellmer:

```{r}
install.packages("ollamar")
install.packages("ellmer")
```

Once these are installed, install kuzco:

```{r}
devtools::install_github("frankiethull/kuzco")
```
