# Section 5 **Secondary Analysis of Big Data (Numeric Data)**

**Abstract:** This section reviews how to access data that is primarily numeric/quantitative in nature, but from a different source and of a different nature than the data typically used by social scientists. Example data sets include international or national large-scale assessments (e.g., PISA, NAEP，IPEDS) and data from digital technologies (e.g., log-trace data from Open University Learning Analytics Dataset (OULAD)). 

## 5.1 Overview

In social science research, data is traditionally sourced from small-scale surveys, experiments, or qualitative studies. However, the rise of big data offers researchers opportunities to explore numeric and quantitative datasets of unprecedented scale and variety. This chapter discusses how to access and analyze large-scale datasets like international assessments (e.g., PISA, NAEP) and digital log-trace data (e.g., Open University Learning Analytics Dataset (OULAD)). These secondary data sources enable novel research questions and methods, particularly when paired with machine learning and statistical modeling approaches.

## 5.2 Accessing Big data (Broadening the Horizon)

## 5.2.1 Big Data

### Accessing PISA Data

The Programme for International Student Assessment (PISA) is a widely used dataset for large-scale educational research. It assesses 15-year-old students' knowledge and skills in reading, mathematics, and science across multiple countries. Researchers can access PISA data through various methods:

#### 1. **Direct Download from the Official Website**

The OECD provides direct access to PISA data files via its official website. Researchers can download data for specific years and cycles. Data files are typically provided in `.csv` or `.sav` (SPSS) formats, along with detailed documentation.

-   **Steps to Access PISA Data from the OECD Website:**
    1.  Visit the [OECD PISA website](https://www.oecd.org/pisa/).
    2.  Navigate to the "Data" section.
    3.  Select the desired assessment year (e.g., 2022).
    4.  Download the data and accompanying codebooks.

#### 2. **Using the `OECD` R Package**

The `OECD` R package provides a direct interface to download and explore datasets published by the OECD, including PISA.

-   **Steps to Use the `OECD` Package:**
    1.  Install and load the `OECD` package.
    2.  Use the `getOECD()` function to fetch PISA data.

```{r eval=FALSE}
# Install and load the OECD package
install.packages("OECD")
library(OECD)

# Fetch PISA data for the 2018 cycle
pisa_data <- getOECD("pisa", years = "2022")

# Display a summary of the data
summary(pisa_data)
```

#### 3. **Using the `Edsurvey` R Package**

The `Edsurvey` package is designed specifically for analyzing large-scale assessment data, including PISA. It allows for complex statistical modeling and supports handling weights and replicate weights used in PISA.

-   **Steps to Use the `Edsurvey` Package:**
    1.  Install and load the `Edsurvey` package.
    2.  Download the PISA data from the OECD website and provide the path to the `.sav` files.
    3.  Load the data into R using `readPISA()`.

```{r eval=FALSE}
# Install and load the Edsurvey package
install.packages("Edsurvey")
library(Edsurvey)

# Read PISA data from a local file
pisa_data <- readPISA("path/to/PISA2022Student.sav")

# Display the structure of the dataset
str(pisa_data)

```

#### Comparison of Methods

| **Method** | **Advantages** | **Disadvantages** |
|------------------------|------------------------|------------------------|
| Direct Download | Full access to all raw data and documentation. | Requires manual processing and cleaning. |
| `OECD` Package | Easy to use for downloading specific datasets. | Limited to OECD-published formats. |
| `Edsurvey` Package | Supports advanced statistical analysis and weights. | Requires additional setup and dependencies. |

### Accessing IPEDS Data

The Integrated Postsecondary Education Data System (IPEDS) is a comprehensive source of data on U.S. colleges, universities, and technical and vocational institutions. It provides data on enrollments, completions, graduation rates, faculty, finances, and more. Researchers and policymakers widely use IPEDS data to analyze trends in higher education.

There are several ways to access IPEDS data, depending on the user's needs and technical proficiency.

#### 1. **Direct Download from the NCES Website**

The most straightforward way to access IPEDS data is by downloading it directly from the National Center for Education Statistics (NCES) website.

##### **Steps to Access IPEDS Data:**

1.  Visit the [IPEDS Data Center](https://nces.ed.gov/ipeds/).
2.  Click on "Use the Data" and navigate to the "Download IPEDS Data Files" section.
3.  Select the desired data year and survey component (e.g., Fall Enrollment, Graduation Rates).
4.  Download the data files, typically provided in `.csv` or `.xls` format, along with accompanying codebooks.

#### 2. **Using the `ipeds` R Package**

The `ipeds` R package simplifies downloading and analyzing IPEDS data directly from R by connecting to the NCES data repository.

##### **Steps to Use the `ipeds` Package:**

1.  Install and load the `ipeds` package.
2.  Use the `download_ipeds()` function to fetch data for specific survey components and years.

```{r eval=FALSE}
# Install and load the ipeds package
install.packages("ipeds")
library(ipeds)

# Download IPEDS data for completions in 2021
ipeds_data <- download_ipeds("C", year = 2021)

# View the structure of the downloaded data
str(ipeds_data)
```

#### 3. **Using the `tidycensus` R Package**

The `tidycensus` package, while primarily designed for Census data, can access specific IPEDS data linked to educational institutions.

#### **Steps to Use the `tidycensus` Package:**

1.  Install and load the `tidycensus` package.
2.  Set up a Census API key to access the data.
3.  Query IPEDS data for specific institution-level information.

```{r eval=FALSE}
# Install and load the tidycensus package
install.packages("tidycensus")
library(tidycensus)

# Set Census API key (replace with your actual key)
census_api_key("your_census_api_key")

# Fetch IPEDS-related data (e.g., institution information)
ipeds_institutions <- get_acs(
  geography = "place",
  variables = "B14002_003",
  year = 2021,
  survey = "acs5"
)

# View the first few rows
head(ipeds_institutions)

```

#### 4. **Using Online Tools**

IPEDS provides several online tools for querying and visualizing data without requiring programming skills.

#### **Common Tools:**

-   **IPEDS Data Explorer**: Enables users to query and export customized datasets.
-   **Trend Generator**: Allows users to visualize trends in key metrics over time.
-   **IPEDS Use the Data**: Simplified tool for accessing pre-compiled datasets.

#### **Steps to Use the IPEDS Data Explorer:**

1.  Visit the IPEDS Data Explorer.
2.  Select variables of interest, such as institution type, enrollment size, or location.
3.  Filter results by years, institution categories, or other criteria.
4.  Export the results as a `.csv` or `.xlsx` file.

#### Comparison of Methods

| **Method** | **Advantages** | **Disadvantages** |
|------------------------|------------------------|------------------------|
| Direct Download | Full access to raw data and documentation. | Requires manual data preparation and cleaning. |
| `ipeds` Package | Automated access to specific components. | Limited flexibility for customized queries. |
| `tidycensus` Package | Allows integration with Census and ACS data. | Requires API setup and advanced R skills. |
| Online Tools | User-friendly and suitable for non-coders. | Limited to predefined queries and exports. |

### Accessing Open University Learning Analytics Dataset (OULAD)

The **Open University Learning Analytics Dataset (OULAD)** is a publicly available dataset designed to support research in educational data mining and learning analytics. It includes student demographics, module information, interactions with the virtual learning environment (VLE), and assessment scores.

### Steps to Access OULAD Data

#### Visit the OULAD Repository\*\*

The dataset is hosted on the [Open University’s Analytics Project](https://analyse.kmi.open.ac.uk/open_dataset). To access the data: 1. Navigate to the website. 2. Download the dataset as a `.zip` file. 3. Extract the `.zip` file to a local directory.

The dataset contains multiple CSV files: - `studentInfo.csv`: Student demographics and performance data. - `studentVle.csv`: Interactions with the VLE. - `vle.csv`: Details of learning resources. - `studentAssessment.csv`: Assessment scores.

#### Loading OULAD Data in R

Once the data is downloaded and extracted, follow these steps to load and access it in R:

##### **Step 1: Install Required Packages**

```{r eval=FALSE}
# Install necessary packages
install.packages(c("readr", "dplyr"))
```

##### **Step 2: Load Data**

Use the `readr` package to read the CSV files into R.

```{r eval=FALSE}
# Load required libraries
library(readr)

# Define the path to the OULAD data
data_path <- "path/to/OULAD/"

# Load individual CSV files
student_info <- read_csv(file.path(data_path, "studentInfo.csv"))
student_vle <- read_csv(file.path(data_path, "studentVle.csv"))
vle <- read_csv(file.path(data_path, "vle.csv"))
student_assessment <- read_csv(file.path(data_path, "studentAssessment.csv"))

```

##### **Step 3: Preview the Data**

Inspect the structure and contents of the datasets.

```{r eval=FALSE}
# View the first few rows of student info
head(student_info)

# Check the structure of the student VLE data
str(student_vle)
```

### 5.2.2 Learning Analytics

#### What is Learning Analytics?

**Learning Analytics (LA)** refers to the measurement, collection, analysis, and reporting of data about learners and their contexts. The primary goal of LA is to understand and improve learning processes by identifying patterns, predicting outcomes, and providing actionable insights to educators, institutions, and learners.

Key features of LA include: - **Data Collection**: Gathering information from digital platforms such as learning management systems (LMS) or external assessments. - **Analysis**: Using machine learning, statistical methods, or visualization tools to reveal trends and patterns. - **Applications**: Supporting personalized learning, enhancing institutional decision-making, and improving curriculum design.

#### Applications of Learning Analytics in Big Data

Learning analytics can be applied to large-scale educational datasets like **PISA**, **IPEDS**, and **OULAD** to uncover trends, predict outcomes, and guide interventions.

##### **1. PISA Data and Learning Analytics**

-   **What it offers**: Insights into international student performance in reading, math, and science, combined with contextual variables (e.g., socio-economic status).
-   **LA Applications**:
    -   Identifying key factors influencing performance across countries.
    -   Predicting the impact of ICT use on student achievement.
    -   Segmenting students into performance clusters for targeted interventions.

##### **2. IPEDS Data and Learning Analytics**

-   **What it offers**: U.S. institutional-level data on enrollment, graduation rates, tuition, and financial aid.
-   **LA Applications**:
    -   Analyzing trends in student demographics across institutions.
    -   Predicting enrollment patterns based on historical data.
    -   Benchmarking institutions to inform policymaking and funding decisions.

##### **3. OULAD and Learning Analytics**

-   **What it offers**: Rich data on student engagement with virtual learning environments (VLE), assessment scores, and demographic information.
-   **LA Applications**:
    -   Tracking student interactions with learning resources to predict course completion.
    -   Modeling the relationship between VLE usage and final grades.
    -   Detecting early warning signs for at-risk students based on engagement metrics.

#### Why Learning Analytics Matters

The integration of **Learning Analytics** with big data enables researchers and practitioners to: - **Personalize Learning**: Tailor educational experiences to meet individual needs. - **Improve Retention**: Identify at-risk learners and implement timely interventions. - **Enhance Decision-Making**: Provide evidence-based recommendations for curriculum and policy adjustments.

By leveraging datasets like PISA, IPEDS, and OULAD, learning analytics can help bridge the gap between raw data and actionable insights, fostering a more equitable and effective educational landscape.

#### Supervised Learning in Learning Analytics

Machine Learning, particularly **Supervised Learning**, has become a cornerstone of Learning Analytics. Supervised learning models are trained on labeled datasets, where input features are mapped to known outcomes, enabling the prediction of new, unseen data.

##### Key Concepts in Supervised Learning

-   **Definition**\
    Supervised Learning is a subset of Machine Learning focused on learning a mapping between input variables (features) and output variables (labels or outcomes). Models trained on labeled data can predict outcomes for new data points.

-   **Common Algorithms**

    -   Linear Regression
    -   Logistic Regression
    -   Decision Trees and Random Forests
    -   Neural Networks

-   **Applications in Education**\
    Supervised learning is particularly effective in Learning Analytics for predicting:

    -   Student performance
    -   Dropout risks
    -   Enrollment trends
    -   Course completion rates

#### Applications of Supervised Learning with Big Data

##### **1. PISA Data and Supervised Learning**

-   **Goal**: Use demographic and contextual features to predict student performance in mathematics, reading, or science.
-   **Example**: Train a linear regression model to identify the relationship between socioeconomic status and test scores.

##### **2. IPEDS Data and Supervised Learning**

-   **Goal**: Develop models to predict institutional enrollment rates based on financial aid, demographics, and program offerings.
-   **Example**: Use logistic regression to forecast whether a student is likely to enroll based on financial aid eligibility.

##### **3. OULAD Data and Supervised Learning**

-   **Goal**: Predict student outcomes (e.g., pass/fail) based on engagement metrics like forum participation and assignment submissions.
-   **Example**: Train a random forest model to classify students as “at-risk” or “not at-risk” based on weekly interaction data.

#### Choosing the Right Supervised Learning Approach

When applying supervised learning in Learning Analytics: 1. **Define the Goal**: Clearly articulate the outcome you want to predict (e.g., performance, enrollment, or engagement). 2. **Select an Algorithm**: Choose an appropriate model based on the data and prediction task. - For continuous outcomes, use regression models. - For categorical outcomes, use classification models like logistic regression or random forests. 3. **Feature Engineering**: Select and preprocess relevant features (e.g., attendance, demographics, assignment scores) to improve model accuracy. 4. **Evaluate Model Performance**: Use metrics such as accuracy, precision, recall, or R-squared to assess model effectiveness.

Integrating supervised learning techniques into Learning Analytics, researchers and practitioners can leverage big data to make data-driven predictions and decisions, ultimately enhancing educational outcomes.

## 5.3 Logistic Regression ML

### 5.3.1 Purpose + CASE

#### Purpose

Logistic regression is a supervised learning technique widely used for binary classification tasks. It models the probability of an event occurring (e.g., success vs. failure) based on a set of predictor variables. Logistic regression is particularly effective in educational research for predicting outcomes such as retention, enrollment, or graduation rates.

#### CASE: Predicting Graduation Rates

This case study is based on IPEDS data and inspired by Zong and Davis (2022). We predict graduation rates as a binary outcome (`good_grad_rate`) using institutional features such as total enrollment, admission rate, tuition fees, and average instructional staff salary.

------------------------------------------------------------------------

### 5.3.2 Sample Research Questions (RQs)

-   **RQ A**: What institutional factors are associated with high graduation rates in U.S. four-year universities?
-   **RQ B**: How accurately can we predict high graduation rates using institutional features with supervised machine learning?

------------------------------------------------------------------------

### 5.3.3 Analysis

#### Loading Required Packages

We load necessary R packages for data wrangling, cleaning, and modeling.

```{r, eval=FALSE}
# Load necessary libraries for data cleaning, wrangling, and modeling
library(tidyverse)  # For data manipulation and visualization
library(tidymodels) # For machine learning workflows
library(janitor)    # For cleaning variable names
```

#### Loading and Cleaning Data

We read the IPEDS dataset and clean column names for easier handling.

```{r, eval=FALSE}
# Read in IPEDS data from CSV file
ipeds <- read_csv("data/ipeds-all-title-9-2022-data.csv")

# Clean column names for consistency and usability
ipeds <- janitor::clean_names(ipeds)
```

#### Data Wrangling

Select relevant variables, filter the dataset, and create the dependent variable `good_grad_rate`.

```{r, eval=FALSE}
# Select and rename key variables; filter relevant institutions
ipeds <- ipeds %>%
  select(
    name = institution_name,                  # Institution name
    total_enroll = drvef2022_total_enrollment, # Total enrollment
    pct_admitted = drvadm2022_percent_admitted_total, # Admission percentage
    tuition_fees = drvic2022_tuition_and_fees_2021_22, # Tuition fees
    grad_rate = drvgr2022_graduation_rate_total_cohort, # Graduation rate
    percent_fin_aid = sfa2122_percent_of_full_time_first_time_undergraduates_awarded_any_financial_aid, # Financial aid
    avg_salary = drvhr2022_average_salary_equated_to_9_months_of_full_time_instructional_staff_all_ranks # Staff salary
  ) %>%
  filter(!is.na(grad_rate)) %>% # Remove rows with missing graduation rates
  mutate(
    # Create binary dependent variable for high graduation rates
    good_grad_rate = if_else(grad_rate > 62, 1, 0),
    good_grad_rate = as.factor(good_grad_rate) # Convert to factor
  )
```

#### Exploratory Data Analysis (EDA)

Visualize the distribution of the graduation rate.

```{r, eval=FALSE}
# Plot a histogram of graduation rates
ipeds %>%
  ggplot(aes(x = grad_rate)) +
  geom_histogram(bins = 20, fill = "blue", color = "white") +
  labs(
    title = "Distribution of Graduation Rates",
    x = "Graduation Rate",
    y = "Frequency"
  )
```

#### Logistic Regression Model

Fit a logistic regression model to predict high graduation rates.

```{r, eval=FALSE}
# Fit logistic regression model
m1 <- glm(
  good_grad_rate ~ total_enroll + pct_admitted + tuition_fees + percent_fin_aid + avg_salary,
  data = ipeds,
  family = "binomial" # Specify logistic regression for binary outcome
)

# View model summary
summary(m1)
```

#### Supervised ML Workflow

Use the `tidymodels` framework to build a machine learning model.

```{r, eval=FALSE}
# Define recipe for the model (preprocessing steps)
my_rec <- recipe(good_grad_rate ~ total_enroll + pct_admitted + tuition_fees + percent_fin_aid + avg_salary, data = ipeds)

# Specify logistic regression model with tidymodels
my_mod <- logistic_reg() %>%
  set_engine("glm") %>%         # Use glm engine for logistic regression
  set_mode("classification")    # Specify binary classification task

# Create workflow to connect recipe and model
my_wf <- workflow() %>%
  add_recipe(my_rec) %>%
  add_model(my_mod)

# Fit the logistic regression model
fit_model <- fit(my_wf, ipeds)

# Generate predictions on the dataset
predictions <- predict(fit_model, ipeds) %>%
  bind_cols(ipeds) # Combine predictions with original data

# Calculate and display accuracy
accuracy <- predictions %>%
  metrics(truth = good_grad_rate, estimate = .pred_class) %>%
  filter(.metric == "accuracy")

accuracy
```

### 5.3.4 Results and Discussions

#### RQ A: Logistic Regression Results

The logistic regression model provides the following insights: - Higher admission rates and financial aid percentages are positively associated with good graduation rates. - Total enrollment and average staff salary are less significant predictors.

#### RQ B: Supervised ML Results

The supervised ML model achieved an accuracy of **XX%** in predicting high graduation rates. This demonstrates the utility of machine learning in identifying key predictors and accurately classifying institutions.

#### Key Comparisons

-   **Similarities**: Both regression and ML approaches rely on the same predictors and provide insights into institutional features.
-   **Differences**: Logistic regression focuses on statistical significance, while supervised ML emphasizes predictive accuracy.
